An Analysis of Massachusetts' Standardized Testing through Multi-Group Structural Equation Modelling

### 2014-03-24 paper-knitr.Rnw
updates3

Missing data in survey and test data is an important consideration, and there exist a number of techniques to produce more robust solutions; most work under the Bayesian estimation process for missing data pioneered by Donald Rubin\cite{rubin}. However, within the structure of the exam, missing data was unsurprisingly a rare occurrence: the highest proportion of missing for any item was .008, and 91.23\% had no missing questions. Even so, for the MG-CFA we attempted to account for this by utilizing the Full Information Maximum Likelihood (FIML) estimation in conjunction with a large number of bootstraps to obtain accurate estimates of asymptotic errors. However, the questionnaire exhibited extreme proportions of non-response, which is problematic for the multivariate regression techniques as it produces potentially biased results under the assumption that items are missing completely at random. As such, we processed the original responses using the mice package\cite{mice} using 13 imputed data frames with 15 iterations in each dataframe. Regression trees are relatively robust to missing data because all information is not processed simultaneously, but rather sequentially in order of variable importance. As such we did not take any additional steps when conducting that analysis.
\section{Results}
\subsection{Exploratory factor analyses}
Given the preceding information, we examined the Mathematics, English, and Biology sections of the exam using a oblimin rotated factor structure, with the number of latent factors equal to the number of general domains. Factor loadings were calculated using maximum likelihood estimates from the tetrachoric correlation matrices for each subset of items. The psych package was used to calculate the factor structure. Items that loaded higher than .3 on a factor were specified as indicators, and diagrams of each factor can be located in Appendix A.

@@

\subsubsection{Multivariate Regression}
A parametric regression model was also conducted exploring only main effects between the predictors and response. Secondary analyses run upon the data, using the VIM package, clearly indicated that the missing data was uniformly equal for all but one question, which substantively improves the likelihood that these questions are truly missing completely at random (MCAR). In addition, the bulk of the missing information was a result of all questions being skipped together. Unfortunately, more algorithmically driven tests \cite{little,KB}, traditionally require that the data in question be normally distributed, and then proceed with homoscedatic assessments of the data; for our purposes, the predictors are all categorical, thus never distributed normally. All tests results were based on both the imputed and missing data, however no differences were found amongst the missing results. Results were singly focused upon the science sections, with more involved educational settings supporting higher outcomes. Males typically outperformed women, and Hispanic/Latino and Black students performed significantly worse than other ethnicity groups, who performed non-significantly differentially from themselves. Otherwise, student achievement outcomes were the only conclusively meaningful predictors that presented, with students who expected to achieve higher (e.g., achieve a baccalaureate degree) having higher outcomes. All code necessary to replicate these effects are submitted within the R script file submitted with these results.

>>>>

Missing data in survey and test data is an important consideration, and there exist a number of techniques to produce more robust solutions; most work under the Bayesian estimation process for missing data pioneered by Donald Rubin\cite{rubin}. However, within the structure of the exam, missing data was unsurprisingly a rare occurrence: the highest proportion of missing for any item was .008, and 91.23\% had no missing questions. Even so, for the MG-CFA we attempted to account for this by utilizing the Full Information Maximum Likelihood (FIML) estimation in conjunction with a large number of bootstraps to obtain accurate estimates of asymptotic errors. However, the questionnaire exhibited extreme proportions of non-response, which is problematic for the multivariate regression techniques as it produces potentially biased results under the assumption that items are missing completely at random. As such, we processed the original responses using the mice package\cite{mice} using 13 imputed data frames with 15 iterations in each dataframe. Regression trees are relatively robust to missing data because all information is not processed simultaneously, but rather sequentially in order of variable importance. As such we did not take any additional steps when conducting that analysis.
=======
Missing data in survey and test data is an important consideration, and there exist a number of techniques to produce more robust solutions; most work under the bayesian estimation process for missing data pioneered by Donald Rubin\cite{rubin}. However, within the structure of the exam, missing data was unsurprisingly a rare occurence: the highest proportion of missing for any item was .008, and 91.23\% had no missing questions. Even so, for the MG-CFA we attempted to account for this by utilising the Full Information Maximum Likelihood (FIML) estimation in conjunction with a large number of bootstraps to obtain accurate estimates of asymptotic errors. However, the questionnaire exhibited extreme proportions of non-response, which is problematic for the multivariate regression techniques as it produces potentially biased results under the assumption that items are missing completely at random. As such, we processed the original responses using the mice package\cite{mice} using 13 imputed data frames with 15 iterations in each dataframe. Regression trees are relatively robust to missing data because all information is not processed simultaneously, but rather sequentially in order of variable importance. As such we did not take any additional steps when conducting that analysis.
>>>>>>> 782b5472c392dad1d398b23760e87d19f21ebe4a
\section{Results}
\subsection{Exploratory factor analyses}
Given the preceding information, we examined the Mathematics, English, and Biology sections of the exam using a oblimin rotated factor structure, with the number of latent factors equal to the number of general domains. Factor loadings were calculated using maximum likelihood estimates from the tetrachoric correlation matrices for each subset of items. The psych package was used to calculate the factor structure. Items that loaded higher than .3 on a factor were specified as indicators, and diagrams of each factor can be located in Appendix A.

@@

\subsubsection{Multivariate Regression}

A parametric regression model was also conducted exploring only main effects between the predictors and response. Secondary analyses run upon the data, using the VIM package, clearly indicated that the missing data was uniformly equal for all but one question, which substantively improves the likelihood that these questions are truly missing completely at random (MCAR). In addition, the bulk of the missing information was a result of all questions being skipped together. Unfortunately, more algorithmically driven tests \cite{little,KB}, traditionally require that the data in question be normally distributed, and then proceed with homoscedatic assessments of the data; for our purposes, the predictors are all categorical, thus never distributed normally. All tests results were based on both the imputed and missing data, however no differences were found amongst the missing results. Results were singly focused upon the science sections, with more involved educational settings supporting higher outcomes. Males typically outperformed women, and Hispanic/Latino and Black students performed significantly worse than other ethnicity groups, who performed non-significantly differentially from themselves. Otherwise, student achievement outcomes were the only conclusively meaningful predictors that presented, with students who expected to achieve higher (e.g., achieve a baccalaureate degree) having higher outcomes. All code necessary to replicate these effects are submitted within the R script file submitted with these results.
